## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical) {
  do 
} else {
  do  else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10
if(a <= 5) {
  print(a)
}else {
  print("it's more than 5")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 10
if(a <= 5) {
  print("small")
} else if (a >5 & a<=10){
  print("medium")
} else if (a>10 | a==7){
  print("large")
} else {
  print("huge")
}  

```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a/2 !=0){
  if(a==5){
    print("a is 5")
  }
}
    
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5){
  print(paste(value, "is greater than", value -1))
}

```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <-  c(1,1,1,1,1,1)
#change the first , third, and sixth elements to 6
for (i in c(1,3,6)){
a_vector[i] <- 6
}
a_vector
```

We can loop over more than just lists:

```{r}
mat <-  matrix(1, nrow=2, ncol =2)

for (row in 1:2){ 
  for (col in 1:2){
    mat[row,col] <- 0
    print(mat)
  }
  }

```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b< 10){
  #increase b by 2
  b <- b+2
  # print b
  print(b)
}

```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # do some stuff
}
```

An example:

```{r}
#start w a vector of temperatures
test_temp <- c(16.55, 26.94, 32.59, 8.58, 28.003)

#function that converts C to F:
c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) + 32
  
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr


```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
test_mat
apply(test_mat, 1,sum)
apply(test_mat, 2,sum)

#margin of 1 sums by row hence 4 5
#margin of 2 sums by col hence 2 3 4
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.55, 26.94, 32.59, 8.58, 28.003, 16.95, 26.64, 38.59, 18.58, 59.003), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}
# 1    #Task
#Given an integral number, determine if it's a square number:
    
    is_square <- function(n){
  if (n < 0) {
    return(FALSE)
  } else {
    if (sqrt(n) %% 1 == 0) {
      return(TRUE)
    } else {
      return(FALSE)
    }
  }
}

# 2   #We have two consecutive integers k1 and k2, k2 = k1 + 1

#We need to calculate the lowest strictly positive integer n, such that: the values nk1 and nk2 have the same digits but in different order.
    
    find_lowest_int <- function(k) {
    k2 <- k + 1
    n <- 9
    while(!identical(sort(as.character(k * n)), sort(as.character(k2 * n)))) {
        n <- n + 1
    }
    return(n)
    }
    
# 3  #Create a function that takes an integer as an argument and returns "Even" for even numbers or "Odd" for odd numbers.  
    
    is_even_or_odd <- function(n) {
  if (n %% 2 == 0) {
    return("Even")
  } else {
    return("Odd")
  }
}

# 4    #We need a function that can transform a number (integer) into a string.

#What ways of achieving this do you know?
    #python first
    
   # def int_to_string(num):
    #return str(num)


    int_to_string <- function(num){
  return(as.character(num))
    }
    
#5    Convert boolean values to strings 'Yes' or 'No'.
    yes_no_function <- function(bool) {
 if (bool) {
    return("Yes")
  } else {
    return("No")
  }
}


    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}
    fizzbuzz_check <- function(x) {
  if ((x %% 3 == 0) & (x %% 5 == 0)) {
    return("fizzbuzz")
  } else if (x %% 3 == 0) {
    return("fizz")
  } else if (x %% 5 == 0) {
    return("buzz")
  } else {
    return(x)
  }
}

fizzbuzz_check(9)


    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}
    # You may feel more comfortable figuring this out in Python first, then translating into R later.
    marioStairs <- function(n) {
  for (i in 1:n) {
    cat(paste(rep("X", i), collapse = ""), "\n")
  }
}
marioStairs(5)
    ```
